Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-06-01 Ben_Manuscripts/stochastic_transport/response_to_reviewers.tex

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{xr}
\usepackage{xcolor}
\usepackage{siunitx}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{scrextend}

\externaldocument[S-]{Supporting_Information}
\externaldocument[M-]{Final_Draft}

\begin{document}

\graphicspath{{./figures/}}

\begin{center}
\textbf{Response to reviewers: Capturing Subdiffusive Solute Dynamics and Predicting
Selectivity in Nanoscale Pores with Time Series Modeling} \\
Authors: Benjamin J. Coscia and Michael R. Shirts
\end{center}

We thank the reviewers for carefully reading over our manuscript and providing
helpful comments. We have taken the suggestions into consideration and made
appropriate revisions to the manuscript document. The comments have been reproduced
in italics and all changes to the text have been documented below. In cases where 
we modified the text we included the original text (denoted by ``Original text:") 
followed by the new text (denoted by ``New text:").

\section*{Response to Reviewer 1}

\begin{enumerate}[label={Comment \theenumi :}, leftmargin=3.9\parindent]

  \item \textit{\textbf{Page 5 line 19:} In this work, we use the output of our MD simulations
        to construct \textbf{two classes} of mathematical models which aim to predict membrane 
        performance while providing quantitative mechanistic insights. \\
        \textbf{Page 5 line 47:} We treat our system in terms of \textbf{two well-known} classes of 
        anomalous subdiffusion: \textbf{fractional Brownian motion} (FBM) subordinate to a 
        \textbf{continuous time random walk (CTRW)}, or subordinated FBM (sFBM) for short. \\ \\
        I am a little confused by these sentences, does the two well-known classes correspond to FBM
        and CTRW, or refer to different classes of modeling approaches used later in the manuscript?}

  %BJC: use a different word?

  Author reply: We thank the reviewer for identifying our potentially confusing choice of words.
  The first sentence (Page 5 line 19) refers more broadly to the two types of stochastic models 
  formulated in this work: the Anomalous Diffusion model and the Markov state-dependent dynamical
  model. The second quoted sentence (Page 5 line 47) refers to the different types of
  motion (FBM and CTRW) that give rise to anomalous diffusion specifically. The paragraph expands
  upon the introduction given in the previous paragraph. We have modified the second quoted 
  sentence as follows to remove ambiguity.

  Original text:
  \begin{quote}
  We treat our system in terms of two well-known classes of anomalous subdiffusion: 
  \end{quote}

  New text:
  \begin{quote}
  In this first approach, we treat our system in terms of two well-known types of molecular
  motion that leads to anomalous subdiffusion: 
  \end{quote}

  \item \textit{\textbf{MD-section} \\
        \textbf{Page 8 and 9} \\
        The authors refer to ref 13 (where these simulations have been performed before), 
        nevertheless I find that the MD section does not provide sufficient amount of details 
        to be self-contained. Could the authors specify the force field, as well as summarize 
        its reliability for this particular system (often simulations involving changed solutes
        through pores require polarizable force fields)? Next, reading this section I would 
        also like to know how well finite size effects are controlled for with this specific 
        membrane and these specific solutes. Lastly, it would be natural to report the size 
        of the system in this section (I found it eventually in the introduction after 
        searching for it throughout the manuscript).}

  Author reply: We thank the reviewer for pushing us to include more details about our 
  molecular simulations in order to boost reproducibility. The length of the manuscript made
  us wary to include information that could be accessed elsewhere, however there are 
  benefits to keep this very important section self-contained.

  We added the following text in order to inform the reader of the system's size:

  New text:
  \begin{quote}
  Each unit cell is contained $\sim$ 62,000 total atoms. % BJC: do you think he want unit cell dimensions? X x Y x Z? 
  \end{quote}

  We added the following text to address comments related to our choice of force field:

  New text:
  \begin{quote}
  We parameterized monomers and solutes using the Generalized Amber Force Field (GAFF) and
  water molecules using the TIP3P water model.~\cite{} We made each choice in order to
  ensure that our approach is accessible and reproducible for a wide audience and so that
  it is easy to study a broad range of solutes and monomer chemistries in high throughput. 
  The main goal of this work is to develop stochastic models which can reproduce solute
  behavior on MD time scales regardless of the force field. In future work, it may be 
  beneficial to employ specialized force fields, like polarizable force fields in 
  order to improve the accuracy of selectivity predictions.
  \end{quote}

  %BJC: not sure if anything should be added to text
  %BJC: I could use some help on justification
  With respect to finite size effects, we have not made any explicit corrections, but
  feel they are not of great concern for this work. While the tails of the monomers from
  different pores interact across periodic boundaries, we are not interested in their
  diffusion. In our previous work, we have shown that monomer motion is negligible on 
  simulation time scales. Our main focus is on the time series of the solutes in the \
  membrane pores. On the 5 $\mu s$ time scales which we simulate we do not see any 
  solutes cross between pores. Therefore, we expect similar solute behavior in larger
  arrays of pores. A study of finite size effects may be interesting in future work, but
  we do not feel it is justified for this study given the computational cost of studying
  larger systems. We also believe it would have little influence on the way in which 
  we fit our models to the time series.

  % BJC: easy to specify force field.
  % BJC: what to say about polarizable force fields? Focus on methods? Maybe throw in some 
  % references to solute transport with polarizable force fields?
  % finite size effects -- ?? sufficiently long pores in z?
  % BJC: no gaurantee polarizable would be good since there aren't any out of the box 
  % polarizable force fields

  \item \textit{\textbf{Figure 5 page 25:} \\
		First, the symbols within the figure are very small and barely readable, and probably not
		according to authors guidelines for figures. Second, the caption, as with the other captions of
		this manuscript, contain a lot of interpretation that normally are placed in the main text. This
		might be a particular style of the authors that has its benefits, however in this specific case the
		caption fails at explaining what exactly this figure is showing. The way I interpret this figure is:
		the symbol on the y-axis marks a general parameter and the boldface symbol marks the
		parameter shown in the histogram. Am I correct? If so, I would strongly recommend modifying
		y-axis and the caption to make this clear.}

		Author reply:
		%BJC: modify figures
		%BJC: cut down on interpretation in all figures

  \item \textit{\textbf{Page 6 line 55}: The second measure of success is based on the qualitative
        comparison between individual realizations of solute trajectories generated ...\\ \\
		Here I am a little curious. Are there no measures which can be used to quantify the similarity of
		the trajectories, such as autocorrelation functions etc, or are such measures inputs for the
		parametrization of the model which cannot be used?}

		Author reply: In short, we do not believe there are any other suitable methods for comparing
		similarity beyond the checks which we show in this work. Autocorrelation is indeed an input 
		to the anomalous diffusion and Markov state-dependent dynamical model, specifically through
		the Hurst parameter which is generated by fitting to autocorrelation of solute jumps. The
		anomalous diffusion model is only parameterized by the distribution of hop length, the 
		distribution of dwell times and the degree of autocorrelation between hops, all of which
		are measured directly from MD. It may be possible to think of other metrics of success
		for the Markov state-dependent dynamical model, but it fails the qualitative test and
		thus does not merit any further checks.

  \item \textit{One question that came to mind reading the paper was: why is the model only in 
  		one dimension and not two (for example using cylindrical coordinates)?}

  		Author reply: We thank the reviewer for their sensible question. The simplest answer
  		to this is that ultimately we are only concerned with transport along the pore axis. Radial
  		solute transport (with respect to the pore center) is effectively bounded and only 
  		contributes to axial MSD to the extent that it influences axial transport. We 
  		do acknowledge that the radial location of solutes has an influence on solute
  		hopping behavior which is why attempted to create the two mode anomalous diffusion model.
		From a more mathematical standpoint, there are two factors which pose challenges with
		increasing dimensionality. First, we have a relatively low amount of data, and therefore
		it is challenging to adequately parameterize behavior as a continuous function of both 
		$r$ and $z$. Second, it is not trivial to simulate fractional motion in multiple dimensions.
		We were fortunate to find an efficient python implementation of FBM and we needed to
		develop our own approximation of 1D fractional L\'{e}vy motion based on literature. 
		We felt there was little to gain by models more complicated than those showcased in our
		work.

		However, we would like to acknowledge that we have a strong desire to incorporate data
		from multiple dimensions. We are actively investigating a new modeling approach which 
		allows us to overcome these issues and to use cylindrical representations with ease.

  \item \textit{\textbf{Page 35 line 12-14}: \\
   		We do not consider the two mode AD model because it has a broken correlation structure \\ \\
        I was a little disappointed to see only the single mode being used. Could the authors 
        elaborate on this point? Exactly how does this broken correlation structure hinder such
        an investigation?}

        Author reply: We are pleased with the reviewer's interest in the two-mode model. While it
        would be technically simple to apply the same mathematical framework to study the long time
        scale performance of the two-mode model, we feel it may be misleading to the reader.
        As described in the text, the predicted MSDs of the two-mode model may indeed show better 
        qualitative behavior than the one-mode model. Specifically, the two-mode model becomes
        linear on long timescales which may be more consistent with our MD results. However, this 
        long time scale linear behavior of the two-mode model has only a mathematical basis and
        no physical basis. As discussed in the text, the one-mode model has persistent curvature
        since hop correlation persists indefinitely for fractional motion. However, for the two-mode
        model, each time a solute transitions between the pore and tail regions, we must generate a
        new FBM process because we lack a proven technique for simulating fractional 
        motion that incorporates time-dependent variation in the width of the hop length 
        distributions.

        Perhaps, most importantly, Section 3.3 is primarily used to demonstrate our mathematical
        framework. Applying it to the two mode model would be trivial. We chose to use the one-mode 
        model to demonstrate because it passes our qualitative and quantitative checks and is
        mathematically sound. As the reviewer points out, it is likely that our approach to 
        modeling the solute time series is not yet suitable to make quantitative predictions of 
        flux and selectivity. But our approach opens the door for applying the same analysis to 
        improved time series modeling techniques.

  \item \textit{\textbf{Page 35 line 20-42:}\\
        A pore-length of 50 nm is a little underwhelming compared to 7 micrometers. Reading the paper
		I do not understand what lies behind this limitation - what hinders the sampling one really would
		like to have when doing such modeling? Also, you write about RAM-requirements scaling more
		than linearly (assuming with L). What exactly lies behind this scaling? Finally, just out of
		curiosity, are there ways of coarse-graining/renormalizing the random walk such that larger time
		and length scales can be reached? That being said, I am impressed by fit against the scaling
	  	law.}

	  	Author reply: We can appreciate why the reviewer has questioned this limitation.
		When simulating fractional motion, each step must be correlated to all of its previous 
		steps. Therefore the computational cost increases with each added step. We show in the paper
		that the mean first passage time across the pores scales with L$^{\alpha}$, and $\alpha$ is
		always $\geq$ 2. As the degree of anti-correlation between hops increases, the scaling
		parameter also increases. Therefore, we need to generate very long trajectories, with 
		correlation between all data points, in order to sufficiently sample the MFPT distribution 
		before fitting them to Equation 9. This is especially challenging with slow-moving solutes
		like acetic acid. We needed to simulate over 2 million timesteps in order to sufficiently
		sample the MFPT over a 50 nm pore. This procedure becomes computationally expensive very
		quickly, and requires ample RAM to store the trajectories as they constructed.

		Given the quality of our fits in Figure 11b, we felt it was not necessary to carry out the
		calculation for longer pores which would likely offer little extra benefit to the insight we
		already gained.

	  	This may be an issue that could be overcome by developing specialized algorithms and perhaps 
	  	by using a compiled computer language like C++.

	  	%BJC: Could reasonably go beyond 50 nm but 7 microns becomes tough
	  	%BJC: building in the correlation structure is computationally expensive. Reference a paper describing how FBM is simulated
	  	%BJC: could benefit from optimized libraries, but python isn't fast enough
	  	%BJC: also need to specify the number of steps to take beforehand
	  	%BJC: truncated correlation functions may allow us to surpass the limitation

  \item \textit{In this paper the models are only evaluated by their ability of achieving 
		  self-consistency with the trajectories used as input. This is conceptually not ideal. 
		  Are there no experimental data or mathematical models which act as a reference for 
		  the models? Please elaborate.}

		  Author reply:
		  % BJC: maybe just give some background on this project overall. Could add a paragraph
		  % to explain this

  \end{enumerate}

\noindent Optional comment to consider:\\
\begin{addmargin}[5.8em]{0em}
	\textit{Having read the paper thoroughly twice, I still struggle with the acronyms of the different random
	walks. Given that this is not a journal of stochastic processes, it is my opinion that the readability
	and accessibility of the manuscript would be greatly improved by writing the full names of the
	random walks in the text while keeping acronyms in figures. \\}

    %BJC: I have no problem with this

\end{addmargin}

\noindent Small typos to consider: \\
\begin{addmargin}[3.3em]{0em}
\begin{itemize}
  \item \textit{Micro in micrometer should be written with upright Greek letter.} \\ \\
  		Author reply:

  \item \textit{d as “dt” in integrals should be non italic.} \\ \\
  		Author reply:

  \item \textit{Page 10 line 28: T should be in symbol form.} \\ \\
  		Author reply:

  \item \textit{Symbol T is used for both time and transition matrix, perhaps use some matrix notation.} \\ \\
  		Author reply:

  \item \textit{The references should be checked thoroughly by the authors. Ref 13, which is arguably
  the most important reference for this paper is missing the volume number.} \\ \\
		Author reply:
\end{itemize}
\end{addmargin}

\section*{Response to Reviewer 2}

\begin{enumerate}[label={Comment \theenumi :}, leftmargin=3.9\parindent]

    \item \textit{ Abstract: here it is difficult to understand what is meant with "dwell times, hop 
    lengths between dwells and correlation between hops", without reading the whole Introduction (and 
    possibly reference 13, which by the way is not completely listed in the References section). The 
    authors should briefly describe the complex transport behavior observed in their previous study of
    these systems (for example, by including information from the Introduction, i.e., the last paragraph
    of page 3 and first paragraph of page 4).}

    Author reply:

    \item \textit{Some of the acronyms were not defined (or perhaps I missed them), for example 'AD' 
    (page 7) and 'MLE' (page 11)}

    Author reply:

    \item \textit{In the Introduction, specifically the last paragraph of page 3 and first paragraph
    of page 4, the authors should refer to Figure 1 to complement their description of the complex solute
    dynamics.}

    Author reply:

    \item \textit{Caption of Figure 7 should refer the reader to Table 2 for completeness}

    Author reply:

\end{enumerate}

%\bibliographystyle{ieeetr}
%\bibliography{transport}

\end{document}

### 2020-06-01 Ben_Manuscripts/hdphmm/Draft.tex
Comments on draft.

     \item This is likely a hopping state that strongly contributes to the solute's MSD. 
     It is short lived and has large fluctuations which often results in very large hops.
     Like State 1, we see this type of behavior near the pore center and near the ends of the tails.

@@

  The distribution of types of hydrogen bonds and associated hydrogen bond lifetimes
  varies by predominant state. States 1 and 2 are dominated by hydrogen bonds to the 
  ester groups at the ends of the monomer tails and, to a smaller extent, the ether 
  linkages between the monomer head group and tails. State 1 methanol molecules that
  donate hydrogen bonds to the tail oxygen atoms exhibit by far the longest hydrogen
  bond lifetimes among all states. Methanol molecules in states 3 and 4 hydrogen 
  bond to all regions of the monomer relatively evenly, which is consistent with our earlier observation that
  states 3 and 4 are representative of behavior seen radially throughout the membrane
  pores. State 3 has the lowest hydrogen bond and sodium association lifetimes which
  is consistent with its short dwell times. These are fleeting interactions since it
  is likely a high variance transitory state. Finally, methanol molecules in state 5
  hydrogen bond almost exclusively to the carboxylate head groups and ether linkages 
  which is consistent with the low (close to the pore center) radial mean of the 
  cluster.

@@

  The size and chemical functionality of the solutes dictates the interactions which
  influence solute transport. In Figure~\ref{fig:hbonds_assoc_summary}, we show that the
  solutes donate hydrogen bonds and associate with sodium ions to varying degrees.
  In fact methanol, by far, interacts the least by these mechanisms. This is likely 
  because methanol is very small and it is easier for it to partition into the tail
  region where there are fewer groups with which it can interact (see Figure~\ref{fig:rdf_summary}).
  Acetic acid, urea and ethylene glycol are slightly larger than methanol and contain 
  multiple polar substituents which stabilize them closer to the pore region.
  These three solutes also participate in significantly more sodium ion association. 
  In our previous work, we showed that sodium ions frequently associate with the 
  carbonyl groups of acetic acid and urea because the carbonyl oxygen atoms are
  unshielded and have a relatively high partial negative charge.~\cite{coscia_chemically_2019}

  \begin{figure}
  \centering
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{hbonds_assoc_summary.pdf}
  \caption{}\label{fig:hbonds_assoc_summary}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{rdf_summary.pdf}
  \caption{}\label{fig:rdf_summary}
  \end{subfigure}
  %BJC: Is it worth putting a bars with the solute MSDs on this plot
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{dwell_time_summary.pdf}
  \caption{}\label{fig:dwell_time_summary}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{density_comparison.pdf}
  \caption{}\label{fig:density_comparison}
  \end{subfigure}
  \caption{(a) Solutes hydrogen bond and associate with sodium ions
  to different degrees. Sometimes, they participate in both interactions at
  once. (b) We histogrammed the radial means of the unclustered states weighted
  by the length of time spent in that state. Most solutes spend the majority of
  their time close to the pore center. Only methanol has a significant density
  far from the pore center. (c) The 95\textsuperscript{th} percentile of dwell
  times, collected by measuring the length of each disjoint segment according 
  to the clustered state sequences, is inversely related to the solute MSDs
  after a 1000 ns time lag.
  (d) The expected dwell time of solutes increases in states with high local 
  number density of surrounding heavy atoms.
  }\label{fig:summaries}
  \end{figure}

  Solutes which experience longer dwell times tend to have lower MSDs. 
  In Figure~\ref{fig:dwell_time_summary}, we plot the 95\textsuperscript{th}
  percentile of dwell times for each solute. The evidence suggests that the 
  long time scale MSDs (we use the MSDs after a 1000 ns time lag in Figure~\ref{fig:dwell_time_summary})
  are inversely related to dwell times. Both acetic acid and urea participate 
  in many hydrogen bonds and sodium ion association interactions and the states
  visited by both solutes have similar distributions of radial means, however
  acetic acid is subject to much longer dwell times, and hence a lower MSD. 
  This may be because acetic acid is a stronger hydrogen bond donor than urea.

>>>>

     \item This is likely a hopping state that strongly contributes to the solute's MSD. 
     It is short lived and has large fluctuations which often results in very large hops.
     Like state 1, we see this type of behavior near the pore center and near the ends of the tails.

@@

  The distribution of types of hydrogen bonds and associated hydrogen bond lifetimes
  varies by predominant state, as seen in more detail in Figure~\ref{fig:hydrogen_bonding}.
  %MRS: what do you think about this? It maks it easier to read.
  \begin{itemize}
  \item States 1 and 2 primarily have hydrogen bonds to the 
  ester groups at the ends of the monomer tails and, to a smaller extent, the ether 
  linkages between the monomer head group and tails. 
  \item State 1 methanol molecules that
  donate hydrogen bonds to the tail oxygen atoms exhibit by far the longest hydrogen
  bond lifetimes among all states. 
  \item Methanol molecules in states 3 and 4 hydrogen 
  bond to all regions of the monomer relatively evenly, which is consistent with our earlier observation that
  states 3 and 4 are representative of behavior seen radially throughout the membrane
  pores. 
  \item State 3 has the lowest hydrogen bond and sodium association lifetimes which
  is consistent with its short dwell times. These are fleeting interactions since it
  is likely a high variance transitory state. 
  \item Finally, methanol molecules in state 5
  hydrogen bond almost exclusively to the carboxylate head groups and ether linkages 
  which is consistent with the low (close to the pore center) radial mean of the 
  cluster.
  \end{itemize}

@@

  The size and chemical functionality of the solutes dictates the interactions which
  influence solute transport. Figure~\ref{fig:hbonds_assoc_summary} shows that the
  solutes donate hydrogen bonds and associate with sodium ions to varying degrees.
  In fact methanol interacts the least by these mechanisms by far. This is likely 
  because methanol is very small and it is easier for it to partition into the tail
  region where there are fewer groups with which it can interact (see Figure~\ref{fig:rdf_summary}).
  Acetic acid, urea and ethylene glycol are slightly larger than methanol and contain 
  multiple polar substituents which stabilize them closer to the pore region.
  These three solutes also participate in significantly more sodium ion association. 
  In our previous work, we showed that sodium ions frequently associate with the 
  carbonyl groups of acetic acid and urea because the carbonyl oxygen atoms are
  unshielded and have a relatively high partial negative charge.~\cite{coscia_chemically_2019}

  \begin{figure}
  \centering
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{hbonds_assoc_summary.pdf}
  \caption{}\label{fig:hbonds_assoc_summary}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{rdf_summary.pdf}
  \caption{}\label{fig:rdf_summary}
  \end{subfigure}
  %BJC: Is it worth putting a bars with the solute MSDs on this plot
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{dwell_time_summary.pdf}
  \caption{}\label{fig:dwell_time_summary}
  \end{subfigure}
  \begin{subfigure}{0.49\textwidth}
  \includegraphics[width=\textwidth]{density_comparison.pdf}
  \caption{}\label{fig:density_comparison}
  \end{subfigure}
  \caption{(a) Solutes hydrogen bond and associate with sodium ions
  to different degrees. Sometimes, they participate in both interactions at
  once. (b) The radial means of the unclustered states weighted
  by the length of time spent in that state. Most solutes spend the majority of
  their time close to the pore center. Only methanol has a significant density
  far from the pore center. (c) The 95\textsuperscript{th} percentile of dwell
  times, collected by measuring the length of each disjoint segment according 
  to the clustered state sequences, is inversely related to the solute MSDs.
  %MRS: this can be explained in the text.
  %after a 1000 ns time lag.
  (d) The expected dwell time of solutes increases in states with high local 
  number density of surrounding heavy atoms.
  }\label{fig:summaries}
  \end{figure}

  Solutes which experience longer dwell times tend to have lower MSDs. 
  In Figure~\ref{fig:dwell_time_summary}, we plot the 95\textsuperscript{th}
  percentile of dwell times for each solute. The evidence suggests that the 
  long time scale MSDs (we use the MSDs after a 1000 ns time lag in Figure~\ref{fig:dwell_time_summary})
  are inversely related to dwell times. Both acetic acid and urea participate 
  in many hydrogen bonds and sodium ion association interactions and the states
  visited by both solutes have similar distributions of radial means, but
  acetic acid has much longer dwell times, and hence a lower MSD. 
  This may be because acetic acid is a stronger hydrogen bond donor than urea.
